\vspace{-3em} \textbf{ Mapping stellar surfaces\\ II: An interpretable Gaussian process model for light curves }

### 2021-01-29 paper2/ms.tex
softened language about spot contrast

The present paper is organized as follows:
we present an overview of the derivation of the GP in \S\ref{sec:main}
and a suite of tests on synthetic data to show the model is calibrated
in \S\ref{sec:calibration}. We discuss our results and the limitations
of our model in \S\ref{sec:discussion} and present straightforward
extensions of the GP, including its application to time-variable
surfaces, in \S\ref{sec:extensions}.
In \S\ref{sec:conclusions} we summarize our results and discuss
topics we will address in future papers in this series.

@@

\section{A Gaussian Process for starspots}
\label{sec:main}

In this section, we present an overview of the computation of our Gaussian
process, which boils down to computing the mean and covariance of the stellar
flux conditioned on certain physical properties of the star and its starspot
distribution. Most of the math is folded into the Appendix for readability;
readers may want to refer to Appendix~\ref{sec:notation} in particular for
a discussion of the notation and conventions we adopt.

\subsection{Spherical harmonics}

@@

Before we dive in, it is useful to introduce the spherical harmonics,
a set of orthogonal functions on the surface of the sphere which we will use
to describe the intensity field on the surface of a star
(Figure~\ref{fig:ylms}). As we will see below,
the spherical harmonics are a particularly convenient basis in which to
describe starspot distributions%
\footnote{There are, of course, drawbacks to using this basis:
    in particular, the spherical harmonics are smooth, continuous functions that
    struggle (at finite degree $l$) to capture high resolution features such as
    small starspots. We discuss this point at length in \S\ref{sec:tinyspots},
    where we show that our model is useful even when applied to stars with spots
    smaller than the effective resolution of the GP.}, as they will allow us to compute
moments of the intensity distribution analytically. Of more immediate concern,
\citet{Luger2019} showed that there is a linear relationship between the
spherical harmonic expansion of a stellar surface and the total disk-integrated
flux $\mathbf{f}$ (i.e., the light curve)
one would observe as the star rotates about a fixed axis.
If the stellar surface intensity is described by a spherical harmonic
coefficient vector $\mathbf{y}$ (up to a certain degree $l_\mathrm{max}$),
the flux is given by
%
\begin{align}
    \label{eq:fAy}
    \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}}(I, P, \mathbf{u}) \, \mathbf{y}
    \quad,
\end{align}
%
where $\mathbf{1}$ is the ones vector and
$\pmb{\mathcal{A}}$ is the \starry
design matrix, a purely linear operator that transforms from the spherical
harmonic basis to the flux basis; it is
a function of the stellar inclination $I$, the stellar
rotation period $P$, and the stellar limb darkening coefficients $\mathbf{u}$
(see Appendix~\ref{sec:starry} for details).

\subsection{Brief overview of Gaussian processes}
\label{sec:gp-intro}
Despite whatever mystique the words ``Gaussian process'' may evoke, a GP
is nothing but a Gaussian distribution in many (formally infinite)
dimensions. Specifically, it is a Gaussian distribution over
\emph{functions} spanning a continuous domain (in our case, the time domain).
Similar to a multivariate Gaussian, which is described by
a $(K \times 1)$ vector $\pmb{\mu}$ characterizing
the mean of the process and a $(K \times K)$
matrix $\pmb{\Sigma}$ characterizing its covariance,
a GP is fully specified by a mean
function $m(t)$ and a kernel function $k(t, t')$.
To say that a random vector-valued variable $\mathbbb{f}$
defined on a $(K \times 1)$ time array $\mathbf{t}$
is ``distributed as a GP'' means that we may write
%
\begin{align}
    \label{eq:fnormal}
    \mathbbb{f} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad,
\end{align}
%
where the elements of the mean and covariance are given by
$\mu_i = m(t_i)$ and $\Sigma_{i,j} = k(t_i, t_j)$, respectively.%
\footnote{%
    In this paper, we will use blackboard font (i.e., $\mathbbb{f}$) to
    denote random variables and serif font (i.e., $\mathbf{f}$) to denote
    particular realizations of those variables.
    See Appendix~\ref{sec:notation} for a detailed explanation of our notation.
}
Because of this relationship to multivariate Gaussians,
GPs are easy to sample from.%
\footnote{Given a 1-d array \texttt{mean} and a 2-d array \texttt{cov} in \Python,
    sampling from the corresponding GP (if it exists)
    can be done in a single line of code by calling
    \texttt{numpy.random.multivariate\_normal(mean, cov)}.}
But the real showstopper is the application of GPs to inference problems.
Multivariate Gaussian distributions have a closed-form (marginal) likelihood
function, so it is easy to compute the probability
of one's data conditioned on a given value of $\pmb{\mu}$ and $\pmb{\Sigma}$
(i.e., the ``likelihood''; see Equation~\ref{eq:log-like} below).
This can in turn be maximized
to infer the optimal values of the model parameters
or used in a
numerical sampling scheme to compute the probability of those parameters
given the data (i.e., the ``posterior'').
Thanks to modern computer architectures, linear algebra packages, and
GP algorithms,
evaluating the GP likelihood may typically be done in a fraction of a second
for a reasonably-sized dataset (i.e., $K \lesssim 10^4$ datapoints).

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbbb{f}$. Rather, a GP
is a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$.
%
In many applications, particularly when modeling stellar light curves,
it is customary to restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \label{eq:kernel}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, typically chosen from a set of
standard functions with desirable smoothness and spectral properties.
\xxx{REMOVE THESE EXAMPLES; NOT NEEDED.}
Among a vast number of kernels are a few that are commonly used
in modeling stellar variability: the exponential squared kernel,
%
\begin{align}
    \label{eq:expsq}
    k_\mathrm{E^2}(\Delta t)           & = \sigma^2 \exp\left(-\frac{\Delta t^2}{2\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the Mat\'ern-3/2 kernel,}
    \label{eq:mat32}
    k_\mathrm{M^\frac{3}{2}}(\Delta t) & = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) \exp\left(-\frac{\sqrt{3}\Delta t}{\tau}\right)
    \quad,                                                                                                                                           \\
    \intertext{the exponential sine squared kernel,}
    \label{eq:expsinsq}
    k_\mathrm{S^2}(\Delta t)           & = \sigma^2 \exp\left[-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)\right]
    \quad,                                                                                                                                           \\
    \intertext{the cosine kernel,}
    k_\mathrm{C}(\Delta t)             & = \sigma^2 \cos\left( \frac{2\pi \Delta t}{P} \right)
    \quad,
\end{align}
%
or any linear combination and/or product of these. In the expressions above,
$\sigma$, $\tau$, $\Gamma$, and $P$ are
the \emph{hyperparameters} of the kernels.
Specifically, $\sigma^2$ is the variance (the value
of the kernel when $t_i = t_j$), $\tau$ is a typical timescale over which
data are correlated, $\Gamma$ is a dimensionless scale parameter, and $P$ is
a period.
Once a kernel has been chosen, these scalar quantities fully
specify the covariance of the Gaussian process. GP-based inference
entails optimizing or sampling over these few quantities, which in many
cases can be done easily and efficiently.
\xxx{BETTER TRANSITION HERE.}

\subsection{Computing the GP}
\label{sec:gp}
%

@@

unit flux.

@@

\subsection{Marginalizing over inclination}
\label{sec:inclination}
%
As we mentioned in the previous section,
the equations for the mean and covariance of our GP
(Equations~\ref{eq:mean_f} and \ref{eq:cov_f}, respectively) are conditioned
on specific values of the stellar and spot properties. To obtain the posterior distribution
for these parameters, we must typically resort to numerical sampling techniques,
which often scale steeply with the number of parameters. It is therefore generally
desirable to keep the total number of parameters small, especially when
employing the GP in an ensemble setting.
In such a setting, we might have light curves from $M$ stars, all of which
we believe to have similar spot properties (perhaps because they have
similar spectral types and rotation periods, for example).
The total number of parameters in our problem is therefore
%
\begin{align}
    N = 4 M + 5
    \quad,
\end{align}
%
since each of the stars will have their own set of 4 stellar properties
(an inclination, a period, and usually two limb darkening coefficients)
but will all share the same 5 spot properties
$\pmb{\theta}_\bullet$ (by assumption).
For a reasonably sized ensemble of $M=100$
stars, we would have to sample over $N = 405$ parameters.
While large, this number is certainly not absurd, especially by modern standards.
However, it does pose
a problem when considering how complex the posterior distribution for the
spot mapping problem can be. In addition to strong nonlinear degeneracies
between some of the parameters (such as the contrast $c$ and the
number of spots $n$), the posterior is often multimodal, especially in the
stellar inclinations. While modern sampling schemes such as Hamiltonian
Monte Carlo and Nested Sampling may in principle be able to deal with these
issues, in practice it can be very difficult to obtain convergence in a
reasonable amount of time.

@@

Fortunately, there is a better way to reduce the number of parameters in
the problem: we can explicitly marginalize over the stellar inclination.
That is, we may write the mean and covariance of our GP as
%
%
\begin{align}
    \label{eq:mu_marg}
    \pmb{\mu}(P, \mathbf{u}, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbbb{f} \, \Big| \, P, \mathbf{u}, \pmb{\theta}_\bullet \Big]
    \nonumber                                                                                                                                                                                                        \\
     & = \mathbf{1} + \mathbf{e}_I
    \\[0.5em]
    \label{eq:cov_marg}
    \pmb{\Sigma}(P, \mathbf{u}, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbbb{f} \, \mathbbb{f}^\top \, \Big| \, P, \mathbf{u}, \pmb{\theta}_\bullet \Big] - \pmb{\mu}(P, \mathbf{u}, \pmb{\theta}_\bullet) \pmb{\mu}^\top(P, \mathbf{u}, \pmb{\theta}_\bullet)
    \nonumber                                                                                                                                                                                                        \\
     & =
    \mathbf{E}_I
    -
    \mathbf{e}_I \,
    \mathbf{e}_I^\top
\end{align}
%
where we define the inclination first moment integral
%
\begin{align}
    \label{eq:eI}
    \mathbf{e}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(\mathbb{I}, P, \mathbf{u}) \,
    \mathrm{E} \Big[ \mathbbb{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
    p(\mathbb{I}) \, \mathrm{d}\mathbb{I}
\end{align}
%
and the inclination second moment integral
%
\begin{align}
    \label{eq:EI}
    \mathbf{E}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(\mathbb{I}, P, \mathbf{u}) \,
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big] \,
    \pmb{\mathcal{A}}^\top(\mathbb{I}, P, \mathbf{u})
    p(\mathbb{I}) \, \mathrm{d}\mathbb{I}
    \quad.
\end{align}
%
The expectations inside the integrals in the expressions for
$\mathbf{e}_I$ and $\mathbf{E}_I$
are given by
Equations~(\ref{eq:exp_y}) and (\ref{eq:exp_yy}), respectively, and
are computed in Appendix~\ref{sec:integrals}.
%
If we are able to perform the integrals in those expressions,
we can \emph{dramatically} reduce the number of
parameters in our ensemble problem.
%
As we show in Appendix~\ref{sec:inclination}, if we assume that stellar
inclinations are distributed isotropically, these integrals
do in fact have closed-form solutions.

@@

\subsection{Normalization correction}
\label{sec:baseline}

@@

\subsection{Summary}
\label{sec:summary}
%
As the computation of the GP relies on many interdependent equations
scattered throughout the previous sections and the Appendix, it
is useful to summarize the procedure for the case where we marginalize
over the inclination (\S\ref{sec:inclination}) and the light curves are
normalized to their means (\S\ref{sec:baseline}), which is likely to be the primary
use case for our algorithm.

@@

While lengthy (and quite tedious), all of the computations described above rely
on equations whose solutions have a closed form.%
%
\footnote{The exception to this is the
    normalization correction (\S\ref{sec:baseline}), which depends on a
    fast-to-evaluate series and thus adds negligible overhead to the computation.}
%
Moreover, most of the terms in the expectation vectors and matrices may
be computed recursively, and many may be pre-computed, as they do not
depend on user inputs.
%
It is therefore possible to evaluate $\tilde{\pmb{\Sigma}}$ in an
efficient manner. In the companion paper \citep{JOSSPaper}, we discuss our
implementation of the algorithm in a user-friendly \Python package.

@@

The surface maps in the figure show dark, compact features of roughly the expected
size and contrast and at the expected latitudes. However, there are important
differences between these maps and what we would obtain by procedurally adding discrete
circular spots to a gridded stellar surface:
%
\begin{enumerate}
    \item \emph{The spots are not circular.} This is most evident in row (c),
          where some spots are distinctly asymmetric.
    \item \emph{There is significant variance in contrast from one spot to another},
          even though our model
          implicitly treats $c$ as constant. Within spots, the contrast is also
          not constant, even though (again) our model implictly treats it as such.
    \item \emph{There is ringing in the background.} This is apparent to some extent in
          row (b), where there are small fluctuations in the brightness at low
          latitudes where no spots are present.
    \item \emph{There aren't exactly 10 (or 20) spots in those maps.} This is
          most obvious in row (c), where only a few large distinct spots,
          plus maybe (?) a few smaller ones, are visible.
    \item \emph{There are bright spots in addition to dark spots.} This may be
          the most glaring issue. We explicitly model spots as being dark, and yet
          there are (almost) just as many bright spots, particularly in row (d).
          While bright spots (such as plages) certainly exist in reality, we did not
          explicitly ask for them here!
\end{enumerate}
%
While these may appear to be critical shortcomings of our model, it is
important to keep in mind that a model consisting of discrete, circular, constant
contrast spots is likely just as far (or perhaps even farther) from the truth. In fact,
points (1) and (2) above suggest our model is more flexible than the discrete
spot model and thus (potentially) better suited to modeling real stellar surfaces.
Points (3), (4), and (5), on the other hand, are more concerning, since they are due,
respectively, to truncation error in the spherical harmonic expansion,
to an intrinsic limitation of our Gaussian approximation,
and to the fact that Gaussian distributions are symmetric about the mean: a positive
deviation is just as likely as a negative deviation of the same magnitude.%
\footnote{
    Even still, the model favors dark spots over bright spots because the GP mean
    itself is lower than unity, in practice making
    positive deviations \emph{from unity} less likely than negative
    deviations. This is why there are usually more dark spots than bright spots in
    the samples shown in the figure.
}
However, as we have argued before, the true power of the GP is in its
applicability to inference problems. In other words, while our GP has some
undesirable features when used as a prior sampling distribution, the real test
of the GP is when it is faced with data in an inference setting.
As long as the data is sufficiently informative, it does not matter that the
prior has finite support for unphysical configurations, as those
will be confidently rejected.

@@

The input parameters for the default run are shown in Table~\ref{tab:synthetic},
and the corresponding light curves in Figure~\ref{fig:calibration_default_data}.
We run the nested sampler as described in the previous section and transform
the posteriors in $a$ and $b$ to posteriors in $\mu_\phi$ and $\sigma_\phi$ via
Equations~(\ref{eq:beta2gauss}), (\ref{eq:muphi}), and (\ref{eq:sigmaphi}).
The results are shown in Figure~\ref{fig:calibration_default_corner}, where
we correctly infer all five parameters within $2-3$ standard deviations.
Posterior distributions for the spot radius $r$, the central spot latitude $\mu_\phi$,
and the spot contrast $c$ are fairly tight, while the distribution for
the latitudinal scatter $\sigma_\phi$ has wider tails and the distribution for
the number of spots $n$ is very poorly constrained. The latter, in particular, is
degenerate with the spot contrast $c$; we discuss this at length in
\citepalias{PaperI}.

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/brightspots.pdf}
        \oscaption{brightspots}{%
            Comparison of GP samples with dark spots (top) and bright
            spots (center) alongside their corresponding light curves
            viewed at $I = 75^\circ$ (bottom). Without knowledge of
            the baseline (\S\ref{sec:baseline}), it is very
            difficult to differentiate between the two from stellar
            light curves.
            \label{fig:brightspots}
        }
    \end{centering}
\end{figure}

>>>>

The present paper is organized as follows:
we present an overview of the derivation of the GP in \S\ref{sec:gp}
and a suite of tests on synthetic data to show the model is calibrated
in \S\ref{sec:calibration}. We discuss our results and the limitations
of our model in \S\ref{sec:discussion} and present straightforward
extensions of the GP, including its application to time-variable
surfaces, in \S\ref{sec:extensions}.
In \S\ref{sec:conclusions} we summarize our results and discuss
topics we will address in future papers in this series.

@@

\section{A Gaussian Process for starspots}
\label{sec:gp}

In this section, we provide a brief overview of Gaussian processes
(\S\ref{sec:gp-intro}) and spherical harmonics (\S\ref{sec:gp-ylms}),
followed by an outline of the derivation of our interpretable GP
(\S\ref{sec:gp-gp}).
This derivation boils down to computing the mean and covariance of the stellar
flux conditioned on certain physical properties of the star and its starspot
distribution.
In \S\ref{sec:gp-inc} and \S\ref{sec:gp-norm} we derive useful extensions
of the model. For convenience, we summarize the results of this entire section in
\S\ref{sec:gp-summary}.
Most of the math is folded into the Appendix for readability;
readers may want to refer to Appendix~\ref{sec:notation} in particular for
a discussion of the notation and conventions we adopt.

\subsection{Brief overview of Gaussian processes}
\label{sec:gp-intro}

Despite whatever mystique the words ``Gaussian process'' may evoke, a GP
is nothing but a Gaussian distribution in many (formally infinite)
dimensions. Specifically, it is a Gaussian distribution over
\emph{functions} spanning a continuous domain (in our case, the time domain).
Similar to a multivariate Gaussian, which is described by
a $(K \times 1)$ vector $\pmb{\mu}$ characterizing
the mean of the process and a $(K \times K)$
matrix $\pmb{\Sigma}$ characterizing its covariance,
a GP is fully specified by a mean
function $m(t)$ and a kernel function $k(t, t')$.
To say that a random vector-valued variable $\mathbbb{f}$
defined on a $(K \times 1)$ time array $\mathbf{t}$
is ``distributed as a GP'' means that we may write
%
\begin{align}
    \label{eq:fnormal}
    \mathbbb{f} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad,
\end{align}
%
where the elements of the mean and covariance are given by
$\mu_i = m(t_i)$ and $\Sigma_{i,j} = k(t_i, t_j)$, respectively.%
\footnote{%
    In this paper, we will use blackboard font (i.e., $\mathbbb{f}$) to
    denote random variables and serif font (i.e., $\mathbf{f}$) to denote
    particular realizations of those variables.
    See Appendix~\ref{sec:notation} for a detailed explanation of our notation.
}
Because of this relationship to multivariate Gaussians,
GPs are easy to sample from.%
\footnote{Given a 1-d array \texttt{mean} and a 2-d array \texttt{cov} in \Python,
    sampling from the corresponding GP (if it exists)
    can be done in a single line of code by calling
    \texttt{numpy.random.multivariate\_normal(mean, cov)}.}
But, as we alluded to earlier, the real showstopper is the application of GPs to inference problems.
Multivariate Gaussian distributions have a closed-form (marginal) likelihood
function, so it is easy to compute the probability
of one's data conditioned on a given value of $\pmb{\mu}$ and $\pmb{\Sigma}$
(i.e., the ``likelihood''; see Equation~\ref{eq:log-like} below).
This can in turn be maximized
to infer the optimal values of the model parameters
or used in a
numerical sampling scheme to compute the probability of those parameters
given the data (i.e., the ``posterior'').
Thanks to modern computer architectures, linear algebra packages, and
GP algorithms,
evaluating the GP likelihood may typically be done in a fraction of a second
for a reasonably-sized dataset (i.e., $K \lesssim 10^4$ datapoints).

Another big advantage of GPs is their flexibility. GPs are often dubbed
a class of ``non-parametric'' models, given that nowhere in the specification
of the GP is there an explicit functional form for $\mathbbb{f}$. Rather, a GP
is a stochastic
process whose draws can in principle take on \emph{any} functional form,
subject, however, to certain smoothness and correlation criteria
of tunable strictness
that are fully
encoded in the covariance $\pmb{\Sigma}$.
%
In many applications, particularly when modeling stellar light curves,
it is customary to restrict the problem by
assuming that the process is \emph{stationary}, such that we may write
%
\begin{align}
    \label{eq:kernel}
    \Sigma_{i,j} & = k(t_i, t_j)
    \nonumber                                    \\
                 & = k(\left| t_i - t_j \right|)
    \nonumber                                    \\
                 & \equiv k(\Delta t)
    \quad.
\end{align}
%
A stationary process is one that is independent of phase (or, in this case,
the actual value of the time $t$); rather, it depends only on the \emph{difference}
between the phases of two data points. The kernel of a stationary process is
therefore a one-dimensional function, typically chosen from a set of
standard functions with desirable smoothness and spectral properties.

The GP we derive in this paper is stationary and admits a representation
as a one-dimensional kernel function. However, as we show in
\S\ref{sec:gp-norm}, the common practice of normalizing stellar light curves to
their mean or median value breaks this stationarity. For this reason,
it is more convenient to derive and present our GP covariance as a
$(K \times K)$ matrix $\pmb{\Sigma}$ and our GP mean as a
$(K \times 1)$ vector $\pmb{\mu}$ for arbitrary $K$ instead of
as a kernel and a mean function. Note, importantly, that these representations
are equivalent given the definitions above.

% Among a vast number of kernels are a few that are commonly used
% in modeling stellar variability: the exponential squared kernel,
% %
% \begin{align}
%     \label{eq:expsq}
%     k_\mathrm{E^2}(\Delta t)           & = \sigma^2 \exp\left(-\frac{\Delta t^2}{2\tau}\right)
%     \quad,                                                                                                                                           \\
%     \intertext{the Mat\'ern-3/2 kernel,}
%     \label{eq:mat32}
%     k_\mathrm{M^\frac{3}{2}}(\Delta t) & = \sigma^2 \left( 1 + \frac{\sqrt{3}\Delta t}{\tau} \right) \exp\left(-\frac{\sqrt{3}\Delta t}{\tau}\right)
%     \quad,                                                                                                                                           \\
%     \intertext{the exponential sine squared kernel,}
%     \label{eq:expsinsq}
%     k_\mathrm{S^2}(\Delta t)           & = \sigma^2 \exp\left[-\Gamma\sin^2\left( \frac{\pi \Delta t}{P} \right)\right]
%     \quad,                                                                                                                                           \\
%     \intertext{the cosine kernel,}
%     k_\mathrm{C}(\Delta t)             & = \sigma^2 \cos\left( \frac{2\pi \Delta t}{P} \right)
%     \quad,
% \end{align}
% %
% or any linear combination and/or product of these. In the expressions above,
% $\sigma$, $\tau$, $\Gamma$, and $P$ are
% the \emph{hyperparameters} of the kernels.
% Specifically, $\sigma^2$ is the variance (the value
% of the kernel when $t_i = t_j$), $\tau$ is a typical timescale over which
% data are correlated, $\Gamma$ is a dimensionless scale parameter, and $P$ is
% a period.
% Once a kernel has been chosen, these scalar quantities fully
% specify the covariance of the Gaussian process. GP-based inference
% entails optimizing or sampling over these few quantities, which in many
% cases can be done easily and efficiently.

\subsection{Spherical harmonics}
\label{sec:gp-ylms}

@@

Before we dive into the computation of our GP, it is useful to introduce the spherical harmonics,
a set of orthogonal functions on the surface of the sphere which we will use
to describe the intensity field on the surface of a star
(Figure~\ref{fig:ylms}). As we will see below,
the spherical harmonics are a particularly convenient basis in which to
describe starspot distributions%
\footnote{There are, of course, drawbacks to using this basis:
    in particular, the spherical harmonics are smooth, continuous functions that
    struggle (at finite degree $l$) to capture high resolution features such as
    small starspots. We discuss this point at length in \S\ref{sec:tinyspots},
    where we show that our model is useful even when applied to stars with spots
    smaller than the effective resolution of the GP.}, as they will allow us to compute
moments of the intensity distribution analytically. Of more immediate concern,
\citet{Luger2019} showed that there is a linear relationship between the
spherical harmonic expansion of a stellar surface and the total disk-integrated
flux $\mathbf{f}$ (i.e., the light curve)
one would observe as the star rotates about a fixed axis.
If the stellar surface intensity is described by a spherical harmonic
coefficient vector $\mathbf{y}$ (up to a certain degree $l_\mathrm{max}$),
the flux is given by
%
\begin{align}
    \label{eq:fAy}
    \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}}(I, P, \mathbf{u}) \, \mathbf{y}
    \quad,
\end{align}
%
where $\mathbf{1}$ is the ones vector and
$\pmb{\mathcal{A}}$ is the \starry
design matrix, a purely linear operator that transforms from the spherical
harmonic basis to the flux basis; it is
a function of the stellar inclination $I$, the stellar
rotation period $P$, and the stellar limb darkening coefficients $\mathbf{u}$
(see Appendix~\ref{sec:starry} for details).

\subsection{Computing the GP}
\label{sec:gp-gp}
%

@@

unit flux.%
\footnote{
    Note, importantly, that these are not the units we observe in! See \citetalias{PaperI}
    and \S\ref{sec:gp-norm} below.

@@

\subsection{Marginalizing over inclination}
\label{sec:gp-inc}
%
As we mentioned in the previous section,
the equations for the mean and covariance of our GP
(Equations~\ref{eq:mean_f} and \ref{eq:cov_f}, respectively) are conditioned
on specific values of the stellar and spot properties. To obtain the posterior distribution
for these parameters, we must typically resort to numerical sampling techniques,
which often scale steeply with the number of parameters. It is therefore generally
desirable to keep the total number of parameters small, especially when
employing the GP in an ensemble setting.
In such a setting, we might have light curves from $M$ stars, all of which
we believe to have similar spot properties (perhaps because they have
similar spectral types and rotation periods, for example).
The total number of parameters in our problem is therefore
%
\begin{align}
    N = 4 M + 5
    \quad,
\end{align}
%
since each of the stars will have their own set of 4 stellar properties
(an inclination, a period, and usually two limb darkening coefficients)
but will all share the same 5 spot properties
$\pmb{\theta}_\bullet$ (by assumption).
For a reasonably sized ensemble of $M=100$
stars, we would have to sample over $N = 405$ parameters.
While large, this number is certainly not absurd, especially by modern standards.
However, it does pose
a problem when considering how complex the posterior distribution for the
spot mapping problem can be. In addition to strong nonlinear degeneracies
between some of the parameters (such as the contrast $c$ and the
number of spots $n$), the posterior is often multimodal, especially in the
stellar inclinations. While modern sampling schemes such as Hamiltonian
Monte Carlo and Nested Sampling may in principle be able to deal with these
issues, in practice it can be very difficult to obtain convergence in a
reasonable amount of time.

@@

Fortunately, there is a better way to reduce the number of parameters in
the problem: we can explicitly marginalize over the stellar inclination.
That is, we may write the mean and covariance of our GP as
%
%
\begin{align}
    \label{eq:mu_marg}
    \pmb{\mu}(P, \mathbf{u}, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbbb{f} \, \Big| \, P, \mathbf{u}, \pmb{\theta}_\bullet \Big]
    \nonumber                                                                                                                                                                                                        \\
     & = \mathbf{1} + \mathbf{e}_I
    \\[0.5em]
    \label{eq:cov_marg}
    \pmb{\Sigma}(P, \mathbf{u}, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbbb{f} \, \mathbbb{f}^\top \, \Big| \, P, \mathbf{u}, \pmb{\theta}_\bullet \Big] - \pmb{\mu}(P, \mathbf{u}, \pmb{\theta}_\bullet) \pmb{\mu}^\top(P, \mathbf{u}, \pmb{\theta}_\bullet)
    \nonumber                                                                                                                                                                                                        \\
     & =
    \mathbf{E}_I
    -
    \mathbf{e}_I \,
    \mathbf{e}_I^\top
\end{align}
%
where we define the inclination first moment integral
%
\begin{align}
    \label{eq:eI}
    \mathbf{e}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(\mathbb{I}, P, \mathbf{u}) \,
    \mathrm{E} \Big[ \mathbbb{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
    p(\mathbb{I}) \, \mathrm{d}\mathbb{I}
\end{align}
%
and the inclination second moment integral
%
\begin{align}
    \label{eq:EI}
    \mathbf{E}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(\mathbb{I}, P, \mathbf{u}) \,
    \mathrm{E} \Big[ \mathbbb{y} \, \mathbbb{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big] \,
    \pmb{\mathcal{A}}^\top(\mathbb{I}, P, \mathbf{u})
    p(\mathbb{I}) \, \mathrm{d}\mathbb{I}
    \quad.
\end{align}
%
The expectations inside the integrals in the expressions for
$\mathbf{e}_I$ and $\mathbf{E}_I$
are given by
Equations~(\ref{eq:exp_y}) and (\ref{eq:exp_yy}), respectively, and
are computed in Appendix~\ref{sec:integrals}.
%
If we are able to perform the integrals in those expressions,
we can \emph{dramatically} reduce the number of
parameters in our ensemble problem.
%
As we show in Appendix~\ref{sec:gp-inc}, if we assume that stellar
inclinations are distributed isotropically, these integrals
do in fact have closed-form solutions.

@@

\subsection{Normalization correction}
\label{sec:gp-norm}

@@

\subsection{Summary}
\label{sec:gp-summary}
%
As the computation of the GP relies on many interdependent equations
scattered throughout the previous sections and the Appendix, it
is useful to summarize the procedure for the case where we marginalize
over the inclination (\S\ref{sec:gp-inc}) and the light curves are
normalized to their means (\S\ref{sec:gp-norm}), which is likely to be the primary
use case for our algorithm.

@@

While lengthy (and quite tedious), all of the computations described above rely
on equations whose solutions have a closed form.%
%
\footnote{The exception to this is the
    normalization correction (\S\ref{sec:gp-norm}), which depends on a
    fast-to-evaluate series and thus adds negligible overhead to the computation.}
%
Moreover, most of the terms in the expectation vectors and matrices may
be computed recursively, and many may be pre-computed, as they do not
depend on user inputs.
%
It is therefore possible to evaluate $\tilde{\pmb{\Sigma}}$ in an
efficient manner. In the companion paper \citep{JOSSPaper}, we discuss our
implementation of the algorithm in a user-friendly \Python package.

@@

The surface maps in the figure show dark, compact features of roughly the expected
size and contrast and at the expected latitudes. However, there are important
differences between these maps and what we would obtain by procedurally adding discrete
circular spots to a gridded stellar surface:
%
\begin{enumerate}
    \item \emph{The spots are not circular.} This is most evident in row (c),
          where some spots are distinctly asymmetric.
    \item \emph{There is significant variance in contrast from one spot to another},
          even though our model
          implicitly treats $c$ as constant. Within spots, the contrast is also
          not constant, even though (again) our model implictly treats it as such.
    \item \emph{There is ringing in the background.} This is apparent to some extent in
          row (b), where there are small fluctuations in the brightness at low
          latitudes where no spots are present.
    \item \emph{There aren't exactly 10 (or 20) spots in those maps.} This is
          most obvious in row (c), where only a few large distinct spots,
          plus maybe a few smaller ones, are visible.
    \item \emph{There are bright spots in addition to dark spots.} This may be
          the most glaring issue. We explicitly model spots as being dark, and yet
          there are (almost) just as many bright spots, particularly in row (d).
          While bright spots (such as plages) certainly exist in reality, we did not
          explicitly ask for them here!
\end{enumerate}
%
While these may appear to be critical shortcomings of our model, it is
important to keep in mind that a model consisting of discrete, circular, constant
contrast spots is likely just as far (or perhaps even farther) from the truth. In fact,
points (1) and (2) above suggest our model is more flexible than the discrete
spot model and thus (potentially) better suited to modeling real stellar surfaces.
Points (3), (4), and (5), on the other hand, are more concerning, since they are due,
respectively, to truncation error in the spherical harmonic expansion,
to an intrinsic limitation of our Gaussian approximation,
and to the fact that Gaussian distributions are symmetric about the mean: a positive
deviation is just as likely as a negative deviation of the same magnitude.%
\footnote{
    Even still, the model favors dark spots over bright spots because the GP mean
    itself is lower than unity, in practice making
    positive deviations \emph{from unity} less likely than negative
    deviations. This is why there are usually more dark spots than bright spots in
    the samples shown in the figure.
}
However, as we have argued before, the true power of the GP is in its
applicability to inference problems. In other words, while our GP has some
undesirable features when used as a prior sampling distribution, the real test
of the GP is when it is faced with data in an inference setting.
As long as the data is sufficiently informative, it does not matter that the
prior has finite support for unphysical configurations, as those
will be confidently rejected.

@@

The input parameters for the default run are shown in Table~\ref{tab:synthetic},
and the corresponding light curves in Figure~\ref{fig:calibration_default_data}.
We run the nested sampler as described in the previous section and transform
the posteriors in $a$ and $b$ to posteriors in $\mu_\phi$ and $\sigma_\phi$ via
Equations~(\ref{eq:beta2gauss}), (\ref{eq:muphi}), and (\ref{eq:sigmaphi}).
The results are shown in Figure~\ref{fig:calibration_default_corner}, where
we correctly infer all five parameters within $2-3$ standard deviations.
Posterior distributions for the spot radius $r$, the central spot latitude $\mu_\phi$,
and the spot contrast $c$ are fairly tight, while the distribution for
the latitudinal scatter $\sigma_\phi$ has wider tails and the distribution for
the number of spots $n$ is very poorly constrained. The latter, in particular, is
degenerate with the spot contrast $c$; we discuss this at length in
\citetalias{PaperI}.

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/brightspots.pdf}
        \oscaption{brightspots}{%
            Comparison of GP samples with dark spots (top) and bright
            spots (center) alongside their corresponding light curves
            viewed at $I = 75^\circ$ (bottom). Without knowledge of
            the baseline (\S\ref{sec:gp-norm}), it is very
            difficult to differentiate between the two from stellar
            light curves.
            \label{fig:brightspots}
        }
    \end{centering}
\end{figure}
